---
title: "About this Project"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---

Welcome to the website for my final project for the fall 2021 PM-566 semester! My name is Jaxon Abercrombie, and I am a fourth year undergraduate student simultaneously earning a bachelor's degree in Health Promotion & Disease Prevention and master's degree in Public Health Data Science. This homepage provides the most important output from the project, and the other pages of this website dive deeper into the questions stated below more deeply. Plots were mainly divided up by the four main regions of California: Superior California, Southern California, Central California, and the Bay Area. More country-specific analysis can be found on other pages of this website for conserving room on the homepage.

$$\\[.1in]$$

### [Background]{.ul}

Needless to say, the ongoing COVID-19 pandemic continues to disrupt daily life. However, with the roll-out of COVID-19 vaccines starting back in December of 2020, certain strides towards normalcy have been made. California's governor, Gavin Newsom, speaks regularly about the success that our state has had with controlling the virus through immunization efforts. While California is certainly diverse and varies in demographic composition and physical environment by region and county, investigating how vaccine uptake has varied by these geographic locations would be fascinating. Especially as new variants of COVID-19 come about, identifying gaps in vaccine roll-out for different demographic groups and discovering which counties require more vaccination efforts overall is crucial. While increases in cases and death evidently come from these new variants, recognizing whether these surges influence vaccination uptake would be interesting to investigate as well. Additionally, whether someone is "Team Pfizer," "Team Moderna," or "Team Johnson & Johnson" has been a popular topic of discussion since roll-out began. Whether there are legitimate differences in a company's vaccine market share is also a topic of interest.

$$\\[.1in]$$

### [Questions of Interest]{.ul}

The overarching, primary question at hand is: *How have COVID-19 vaccination rates varied by county in California since their initial roll-out?* Furthermore, there are two main secondary questions that dig deeper into the data sets used:

1.  How do trends in cases and deaths potentially affect immunization rates for California as a whole?

2.  How do vaccination efforts vary by vaccine company (Pfizer, Moderna, Johnson & Johnson)?

$$\\[.1in]$$

### [About the Datasets]{.ul}

There were three different sets of data used for this project: (1) data for CA counties and their administered vaccine doses, (2) data regarding COVID-19 cases and deaths for each CA county, and (3) demographic data based on vaccine administration.

The data about vaccine doses for each county, titled "Vaccines by California County," came from LA city's data site at [this link](https://data.lacity.org/COVID-19/Vaccines-by-California-County/rpp7-mevy). This particular set has data from the start of vaccine roll-out in mid-December 2020 up until early November 2021 and includes dose data for each vaccine company per county, county population count, and administration date.

Data regarding COVID-related cases and deaths, titled "Statewide COVID-19 Cases Deaths Tests," came from the California Department of Health and Human Services at [this link](https://data.chhs.ca.gov/dataset/covid-19-time-series-metrics-by-county-and-state/resource/046cdd2b-31e5-4d34-9ed3-b48cdbc4be7a). COVID-related cases and deaths, both raw and cumulative, are included for every day since February 2020 to early November 2021 among many other variables.

The set of data containing information about administered vaccines by demographic characteristics, titled "Vaccine demographic data," was sourced from [this link](https://data.lacity.org/COVID-19/Vaccine-demographic-data/iv7a-6rrq). It contains data about who is fully and partially vaccinated based on age and race and ethnicity.

$$\\[.1in]$$

```{r libraries&sources, include = FALSE}
library(tidyverse)
library(plotly)
library(cowplot)
library(RCurl)
library(scales)

source("datawrangle.R")
```

```{r dataprepare, echo = FALSE, include = FALSE}
# Read in the CSV files
download1 <- getURL("https://data.lacity.org/api/views/rpp7-mevy/rows.csv?accessType=DOWNLOAD")
vaxCA <- read.csv(text = download1)

covidCA <- read.csv(url("https://data.chhs.ca.gov/dataset/f333528b-4d38-4814-bebb-12db1f10f535/resource/046cdd2b-31e5-4d34-9ed3-b48cdbc4be7a/download/covid19cases_test.csv"))

# Merge the first two datasets
covidCA <-
  covidCA %>%
  mutate(county = area)

vaxCA <-
  vaxCA %>%
  mutate(date = as.Date(vaxCA$date, "%m/%d/%Y"))

mergedCA <- merge(vaxCA, covidCA, 
                  by = c("date","county"),
                  all.x = TRUE)

mergedCA <-
  mergedCA %>%
  filter(date >= as.Date("2020-12-15")) %>%
  filter(date <= as.Date("2021-11-03")) %>%
  group_by(date, county) %>% 
  summarise(across(c(total_doses,cumulative_total_doses,
                     pfizer_doses,cumulative_pfizer_doses,
                     moderna_doses,cumulative_moderna_doses,
                     jj_doses,cumulative_jj_doses,
                     partially_vaccinated,
                     total_partially_vaccinated,fully_vaccinated,
                     cumulative_fully_vaccinated,at_least_one_dose,
                     cumulative_at_least_one_dose,population,
                     cases,cumulative_cases,deaths,
                     cumulative_deaths,total_tests,
                     cumulative_total_tests,positive_tests,
                     cumulative_positive_tests), mean, .groups = date))

# Create other variables
mergedCA <-
  mergedCA %>%
  mutate(dose_standard = (cumulative_total_doses/population),
         pfizer_perc = ((cumulative_pfizer_doses/cumulative_total_doses)*100),
         moderna_perc = ((cumulative_moderna_doses/cumulative_total_doses)*100),
         jj_perc = ((cumulative_jj_doses/cumulative_total_doses)*100),
         perc_vaccinated = (cumulative_fully_vaccinated/population)*100,
         perc_partial = (cumulative_at_least_one_dose/population)*100)

superior <- c("Butte", "Colusa", "Del Norte", "Glenn", "Humboldt", "Lake", "Lassen", "Mendocino", "Modoc", "Nevada", "Plumas", "Shasta", "Sierra", "Siskiyou", "Tehama", "Trinity")

central <- c("Alpine", "Amador", "Calaveras", "El Dorado", "Fresno", "Inyo", "Kings", "Madera", "Mariposa", "Merced", "Mono", "Placer", "Sacramento", "San Joaquin", "Stanislaus", "Sutter", "Yuba", "Tulare", "Tuolumne", "Yolo")

bay <- c("Alameda", "Contra Costa", "Marin", "Monterey", "Napa", "San Benito", "San Francisco", "San Mateo", "Santa Clara", "Santa Cruz", "Solano", "Sonoma")

southern <- c("Imperial", "Kern", "Orange", "Riverside", "San Bernardino", "San Diego", "San Luis Obispo", "Santa Barbara", "Ventura", "Los Angeles")
```

$$\\[.1in]$$

### [Figure 1: Vaccination Over Time]{.ul}

```{r fig1, echo = FALSE, warning = FALSE}
mergedCA <- 
  mergedCA %>% 
  mutate(region = case_when(county %in% superior ~ "Superior California",
                            county %in% central ~ "Central California",
                            county %in% bay ~ "Bay Area",
                            county %in% southern ~ "Southern California"))

fig1 <- 
    mergedCA %>%
    ggplot() + 
    geom_line(aes(x = date, 
                  y = cumulative_fully_vaccinated/population*100, 
                  color = county)) +
    labs(title = "Percent Fully Vaccinated Since Roll-out by Region", x = "Date", 
         y = "", color = "County") +
  facet_wrap(~region)

ggplotly(fig1) %>% 
  layout(yaxis = list(title = ""),
             margin = list(l = 100), 
             annotations = c(list(text = "Percent Fully Vaccinated (%)",
                                  x = -0.10,
                                  xref = "paper",
                                  y = 0.5,
                                  showarrow = F,
                                  textangle = -90)))

```

**Description:**

**Findings:**

The figure above depicts how vaccine efforts have been across four main regions of California, with country-specific data illustrated for each region. Feel free to toggle with the available counties to make direct comparisons or remove unwanted lines! The data spans from the debut of the COVID-19 vaccine on December 15th, 2020 up until November 3rd, 2021 (as with all following figures). Each region has a comparable number of counties, making these groupings representative and more reliable for drawing conclusions.Generally, each region experienced similar uptake in vaccination, as seen with almost congruent plot shapes. Additionally, each region and respective county followed a similar path of surging after April 2021's eligibility expansion and began to slow quickly during summer of 2021. However, there are certainly some regional differences, like the Bay Area having the highest percent of fully vaccinated individuals and Superior California the lowest. Furthermore, regions vary in range of the percent of fully vaccinated individuals; the Bay Area has a range of about 20% between its lowest and highest performing counties, while Superior California has a difference of about 30% between its lowest and highest. Overall, it appears that urban areas (Bay Area and Southern California) have more vaccine uptake than rural areas (Central and Superior California) on average, which will be something to consider with subsequent figures on this website. Access comes to mind as an immediate hurdle for vaccination, as physical distance from urban centers may make rural living disadvantageous.

$$\\[.1in]$$

### [Fig 2: Doses, Cases, and Deaths]{.ul}

```{r fig2, echo = FALSE, warning = FALSE}
fig2 <- 
  mergedCA %>%
  ggplot(aes(group = region)) +
  geom_line(mapping = aes(x = date, y = total_doses, color = "Doses Administered")) +
  geom_line(mapping = aes(x = date, y = cases, color = "Case Count")) +
  geom_line(mapping = aes(x = date, y = deaths, color = "Death Count")) +
  labs(title = "Doses, Cases, and Deaths by Region", x = "Date", y = "") +
  scale_colour_manual(values = c("Doses Administered"="#C78888", 
                                   "Case Count"="#88C7C2", 
                                   "Death Count"="#E7D366")) +
  facet_wrap(~region, scales = "free_y", shrink = TRUE) +
  scale_y_continuous(labels = comma) +
  theme(legend.title = element_blank(), plot.title.position = "middle")
  
ggplotly(fig2, tooltip = c("x", "y")) %>% 
  layout(yaxis = list(title = ""),
             margin = list(l = 100), 
             annotations = c(list(text = "Count",
                                  x = -0.15,
                                  xref = "paper",
                                  y = 0.5,
                                  showarrow = F,
                                  textangle = -90)))
```

```{r test, echo = FALSE}
mergedCA[is.na(mergedCA)] = 0
```

Figure 2 illustrates a potential relationship between vaccine uptake and cases and deaths. At first, individuals rushed to be vaccinated for their own safety, but some individuals remained hesitant and lacked motivation to become vaccinated for many months. Though, changes in cases and deaths due to variants made for rises in the number of vaccine doses administered. A rise in cases appeared to motivate greater vaccine uptake after the start of July 2021, likely because of the once novel Delta variant. Compared to the quasi-plateaus of vaccine uptake as seen in Figure 1, there are visible surges between July and October of 2021 for three regions, excluding the Bay Area. This relationship may also have been influenced by policy efforts to require vaccines in public spaces, eligibility for booster shots, and the realization of the effectiveness of vaccination. Nonetheless, it seems natural for people to protect themselves when a threat is imminent, despite prior sentiments against vaccination.

$$\\[.1in]$$

### [Fig 3: Vaccination by Company]{.ul}

```{r fig3, echo = FALSE, warnings = FALSE, message=FALSE}
mergedCA3 <-
  mergedCA %>%
  group_by(region,date) %>%
  summarise(county = county,
            region = region,
            ave_moderna = ave(moderna_perc),
            ave_pfizer = ave(pfizer_perc),
            ave_jj = ave(jj_perc))

mergedCA3[is.na(mergedCA3)] = 0

fig3 <- 
  mergedCA3 %>%
  ggplot(aes(group = region)) +
  geom_line(aes(x = date, y = ave_moderna, color = "Moderna")) +
  geom_line(aes(x = date, y = ave_pfizer, color = "Pfizer")) +
  geom_line(aes(x = date, y = ave_jj, color = "J&J")) +
  scale_colour_manual(values = c("Moderna"="#C78888", 
                                   "Pfizer"="#88C7C2", 
                                   "J&J"="#E7D366")) +
    labs(title = "Vaccination Efforts by Company", 
         x = "Date", y = "", color = "Company") +
    theme(legend.title = element_blank(), 
          axis.title.y = element_text(hjust = -1.5)) +
    facet_wrap(~region)

ggplotly(fig3, tooltip = c("x", "y")) %>% 
  layout(yaxis = list(title = ""),
             margin = list(l = 100), 
             annotations = c(list(
               text = "Percent of Total Doses Administered (%)",
               x = -0.2,
               xref = "paper",
               y = 0.5,
               showarrow = F,
               textangle = -90)))
```

Understanding how vaccination efforts differ by company may influence future roll-outs of vaccines, since timeliness of roll-out and targeting specific areas may make one company a preferable candidate compared to another. Evident in the figure above, the more "rural" regions tend to receive Moderna doses rather than the Pfizer doses that more "urban" regions tend to receive. Johnson & Johnson doses are limited across all regions, and even when accounting for J&J being single-dose, its impact appears limited. That could potentially be because of defects and withdrawal from markets at times. Assuming that populous urban areas were first to receive vaccines because of supply chain ease with product shipment at ports, Pfizer certainly made an impact. Being the first to reach the vaccine market and be prevalent in populous areas made for a sustainable difference in uptake. Despite Moderna's later debut as a COVID-19 vaccine company, its use surged quickly and ultimately remained dominant in Superior and Central California. Whether it was purposefully more available in those two regions is unknown, but its popularity may have been influenced by Pfizer's presence elsewhere. What appears most fascinating about the relationship between Moderna and Pfizer is that Moderna surpassed Pfizer at one point in each region, which may demonstrate a supply surge that lead to temporary or sustained success.

$$\\[.1in]$$

### [Fig 4: Vaccination by Age]{.ul}

```{r fig4, echo = FALSE, warnings = FALSE, message=FALSE}
dataAge <- datawrangle("Age Group")

fig4palette <- c("#8AA399","#7D84B2","#FD96A9","#FFC857")

fig4 <-
  dataAge %>%
  ggplot(mapping = aes(x = demographic_value, y = region_fully_vax, 
                       fill = demographic_value)) +
  geom_bar(stat = "identity") +
  labs(x = "Age Group", y = "", 
       title = "Vaccination Efforts by Age Group", 
       fill = "Age Group") +
  scale_fill_manual(values = fig4palette) +
  facet_wrap(~region)

ggplotly(fig4, tooltip = c("x", "y")) %>% 
  layout(yaxis = list(title = ""),
             margin = list(l = 100), 
             annotations = c(list(
               text = "Percent Fully Vaccinated (%)",
               x = -0.1,
               xref = "paper",
               y = 0.5,
               showarrow = F,
               textangle = -90)))
```

$$\\[.1in]$$

### [Fig 5: Vaccination by Race/Ethnicity]{.ul}

```{r fig5, echo = FALSE, warnings = FALSE, message=FALSE}
dataRace <- datawrangle("Race/Ethnicity")

# Change data error of over 100%
dataRace[6,3] <- 100.00

fig5palette <- c("#C78888", "#88C7C2", "#E7D366", "#8AA399", 
                 "#7D84B2", "#FD96A9", "#FAA04C")

fig5 <-
  dataRace %>%
  ggplot(mapping = aes(x = demographic_value, y = region_fully_vax, 
                       fill = demographic_value)) +
  geom_bar(stat="identity") +
  labs(x = "Race/Ethnicity", y = "", 
       title = "Vaccination Efforts by Age", 
       fill = "Race/Ethnicity") +
  scale_fill_manual(values = fig5palette) +
  theme(axis.text.x = element_blank()) +
  facet_wrap(~region)

ggplotly(fig5, tooltip = c("x", "y")) %>% 
  layout(yaxis = list(title = ""),
             margin = list(l = 100), 
             annotations = c(list(
               text = "Percent Fully Vaccinated (%)",
               x = -0.2,
               xref = "paper",
               y = 0.5,
               showarrow = F,
               textangle = -90)))
```

$$\\[.1in]$$

### [Overall Takeaways]{.ul}

Text

$$\\[.1in]$$

### [Looking Forward]{.ul}

Traditionally, COVID-19 data is shown aggregated across the U.S. "Three-thousand new cases per day" holds less potential to drive interventions than "only 53% of Black people in the Bay Area are fully vaccinated." This project and the creation of interactive plots encouraged me to investigate a variety of social determinants of health like location, age, and race/ethnicity in depth, which is crucial for my future in championing health equity. Moving forward, I am more informed about how California functions beyond the state-level with COVID-19, and I now can inform peers about inequities found from this project. A future version of this project with information about booster shots will be incredibly exciting, since uptake of booster shots may be vastly different than the initial doses of the COVID-19 vaccine. Now equipped with website and visualization skills, I feel eager to take on a new project!

$$\\[.1in]$$

### [Download Full Report]{.ul}

If you want to download the PDF of my entire report for this project, click [here](https://github.com/jaxabercrombie/finalproject-pm566/raw/master/PM-566%20Final%20Project%20Write-up.pdf)!
